---
title: "Student Dropout Prediction"
subtitle: "MATH 4322 - Group 4 Final Report (Logistic Regression)"
author: "Sriharsha Challangi, Kelvin Mai, Thanh Trung Do, Leah Padilla, Tien Hoang"
date: "Spring 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5)
```

\newpage

# Introduction

## Data Description (Thanh Trung Do, Leah Padilla)

The dataset is the **Student Dropout Prediction Dataset** from Kaggle (Ali, 2024), containing 10,000 student records with 18 predictor variables and 1 binary response variable.

**Response variable:**

- `Dropout`: Binary (0 = retained, 1 = dropped out)

**Predictor variables (inputs):**

- *Demographic*: Age, Gender (Male/Female), Family_Income, Parental_Education (None, High School, Bachelor, Master, PhD)
- *Academic*: GPA, Semester_GPA, CGPA, Semester (Year 1--4), Department (Science, Arts, Business, CS, Engineering)
- *Behavioral*: Study_Hours_per_Day, Attendance_Rate, Assignment_Delay_Days, Travel_Time_Minutes, Stress_Index (1--10)
- *Socioeconomic*: Internet_Access (Yes/No), Part_Time_Job (Yes/No), Scholarship (Yes/No)

Three variables (`Family_Income`, `Study_Hours_per_Day`, `Stress_Index`) have 500 missing values each (5%), which we imputed with their respective medians.

## Research Question

**Which factors affect whether a student drops out, and can we accurately predict student dropout?**

The response variable is `Dropout` (binary: 0 or 1), making this a classification problem.

\newpage

# Methods

## Logistic Regression (Thanh Trung Do, Leah Padilla)

### Why Logistic Regression?

Logistic regression is appropriate because our response variable is binary (Dropout = 0 or 1). As discussed in Lecture 8, logistic regression models the probability that $Y$ belongs to one of two categories, and the output is always between 0 and 1 (unlike linear regression, which can predict values outside this range).

**Advantages:**

- Provides interpretable coefficients: each $\hat{\beta}_j$ tells us the change in log-odds per unit increase in $X_j$.
- Outputs probabilities, allowing flexible threshold selection.
- Computationally efficient and well-suited for inference (significance testing of predictors).

**Disadvantages:**

- Assumes a linear relationship between predictors and log-odds, which may not capture complex nonlinear patterns.
- Can be unstable when classes are well-separated (Lecture 10).
- May underperform compared to tree-based methods when interactions between predictors are important.

### Model Formula

The logistic regression model is:

$$\log\left(\frac{P(\text{Dropout} = 1 \mid X)}{1 - P(\text{Dropout} = 1 \mid X)}\right) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_p X_p$$

where the left-hand side is the log-odds (logit), and $X_1, \ldots, X_p$ are the 18 predictor variables. The predicted probability is:

$$P(\text{Dropout} = 1 \mid X) = \frac{\exp(\beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p)}{1 + \exp(\beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p)}$$

We estimate $\beta_0, \beta_1, \ldots, \beta_p$ using maximum likelihood. In R: `glm(Dropout ~ ., family = binomial)`.

### Thought Process for Model Fitting

After fitting the full model with all 18 predictors, we examine the p-values from the `summary()` output. Predictors with large p-values are not statistically significant and may be excluded. We also use **stepwise selection (AIC)** to find a more parsimonious model and compare it against the full model.

### Data Preprocessing

```{r}
library(ggplot2)
library(dplyr)
library(corrplot)

data <- read.csv("student_dropout_dataset_v3.csv")

data$Family_Income[is.na(data$Family_Income)] <- median(data$Family_Income, na.rm = TRUE)
data$Study_Hours_per_Day[is.na(data$Study_Hours_per_Day)] <- median(data$Study_Hours_per_Day, na.rm = TRUE)
data$Stress_Index[is.na(data$Stress_Index)] <- median(data$Stress_Index, na.rm = TRUE)

data$Student_ID <- NULL
data$Gender <- as.factor(data$Gender)
data$Internet_Access <- as.factor(data$Internet_Access)
data$Part_Time_Job <- as.factor(data$Part_Time_Job)
data$Scholarship <- as.factor(data$Scholarship)
data$Semester <- as.factor(data$Semester)
data$Department <- as.factor(data$Department)
data$Parental_Education <- as.factor(data$Parental_Education)
data$Dropout <- as.factor(data$Dropout)
```

```{r}
cat("Dropout distribution:\n")
table(data$Dropout)
cat("\nDropout rate:", round(mean(data$Dropout == "1") * 100, 2), "%\n")
```

### Train/Test Subdivision (80/20, repeated 10 times)

We randomly subdivide the data into 80% training and 20% testing. We repeat this 10 times using `set.seed()` and record the test prediction error (misclassification rate) for each iteration.

```{r}
seeds <- 1:10
n <- nrow(data)

resample_results <- data.frame(
  Iteration = integer(),
  Error_Rate = numeric(),
  Accuracy = numeric(),
  Sensitivity = numeric(),
  Specificity = numeric()
)

for (s in seeds) {
  set.seed(s)
  idx <- sample(1:n, 0.8 * n)
  tr <- data[idx, ]
  te <- data[-idx, ]

  model <- glm(Dropout ~ ., data = tr, family = binomial)
  prob <- predict(model, newdata = te, type = "response")
  pred <- ifelse(prob > 0.5, 1, 0)
  pred <- factor(pred, levels = c(0, 1))

  cm <- table(Predicted = pred, Actual = te$Dropout)
  tp <- cm["1", "1"]; tn <- cm["0", "0"]
  fp <- cm["1", "0"]; fn <- cm["0", "1"]

  resample_results <- rbind(resample_results, data.frame(
    Iteration = s,
    Error_Rate = 1 - (tp + tn) / sum(cm),
    Accuracy = (tp + tn) / sum(cm),
    Sensitivity = tp / (tp + fn),
    Specificity = tn / (tn + fp)
  ))
}

rownames(resample_results) <- NULL
round(resample_results, 4)
```

```{r}
cat("Mean test prediction error across 10 iterations:",
    round(mean(resample_results$Error_Rate), 4), "\n")
```

\newpage

# Results (Thanh Trung Do, Leah Padilla)

## Model Summary

We fit the logistic regression on the full dataset to examine which predictors are significant.

```{r}
data_num <- data
data_num$Dropout <- as.numeric(data_num$Dropout) - 1
fit.full <- glm(Dropout ~ ., data = data_num, family = binomial)
summary(fit.full)
```

## Stepwise Model Selection

```{r}
step_model <- step(fit.full, direction = "both", trace = 0)
cat("Full model AIC: ", AIC(fit.full), "\n")
cat("Stepwise model AIC:", AIC(step_model), "\n")
```

```{r}
anova(step_model, fit.full, test = "Chisq")
```

## Test Prediction Error

```{r, fig.width=7, fig.height=4}
plot(resample_results$Iteration, resample_results$Error_Rate,
     type = "b", pch = 19, col = "tomato",
     xlab = "Iteration", ylab = "Test Error Rate",
     main = "Test Error Rate Across 10 Random 80/20 Subdivisions")
abline(h = mean(resample_results$Error_Rate), lty = 2, col = "blue")
legend("topright",
       legend = paste("Mean =", round(mean(resample_results$Error_Rate), 4)),
       lty = 2, col = "blue")
```

```{r}
cat("=== Logistic Regression - Prediction Error Summary ===\n\n")
cat("Mean test error (10 iterations):", round(mean(resample_results$Error_Rate), 4), "\n")
cat("SD of test error:               ", round(sd(resample_results$Error_Rate), 4), "\n")
cat("Mean accuracy:                  ", round(mean(resample_results$Accuracy), 4), "\n")
```

## Variable Importance

```{r, fig.height=6}
num_vars <- c("Age", "Family_Income", "Study_Hours_per_Day", "Attendance_Rate",
              "Assignment_Delay_Days", "Travel_Time_Minutes", "Stress_Index",
              "GPA", "Semester_GPA", "CGPA")
train_scaled <- data_num
for (col in num_vars) {
  mu <- mean(data_num[[col]]); s <- sd(data_num[[col]])
  train_scaled[[col]] <- (data_num[[col]] - mu) / s
}
model_scaled <- glm(Dropout ~ ., data = train_scaled, family = binomial)
std_coefs <- sort(abs(coef(model_scaled)[-1]), decreasing = TRUE)
top_n <- min(15, length(std_coefs))
par(mar = c(5, 10, 3, 2))
barplot(std_coefs[top_n:1], horiz = TRUE, las = 1,
        col = "steelblue", border = NA,
        main = "Top Predictors (|Standardized Coefficients|)",
        xlab = "|Standardized Coefficient|")
```

## Interpretation

- **Academic performance** variables (GPA, CGPA, Semester_GPA) are among the strongest predictors. Students with lower GPAs have significantly higher odds of dropping out.
- **Behavioral factors** such as Attendance_Rate and Study_Hours_per_Day are also significant.
- The model achieves a **mean test error rate** of `r round(mean(resample_results$Error_Rate), 4)` across 10 random 80/20 subdivisions, with low standard deviation (`r round(sd(resample_results$Error_Rate), 4)`), confirming stable predictive performance.

\newpage

# Conclusion

- Logistic regression achieves a mean test prediction error of **`r round(mean(resample_results$Error_Rate), 4)`** across 10 random subdivisions (80/20).
- The significant predictors (GPA, Attendance_Rate, Study_Hours_per_Day, etc.) align with domain knowledge about student success factors.
- The low standard deviation across iterations confirms model stability.

*Note: The Decision Tree results and final model comparison will be added by the Decision Tree team (Sriharsha, Kelvin, Tien).*

\newpage

# References

1. Ali, M. (2024). Student Dropout Prediction Dataset. Kaggle. https://www.kaggle.com/datasets/meharshanali/student-dropout-prediction-dataset

2. James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning with Applications in R* (2nd ed.). Springer.

3. R Core Team (2025). R: A Language and Environment for Statistical Computing. https://www.R-project.org/

\newpage

# Appendix: Full R Source Code

*The full R source code is attached separately.*
